Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging

### 2023-01-12 script/manuscript_MEE.Rmd
correlation between grains counted and sequence reads classified

---
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging
  Patterns of Queen Bumble Bees
author:  |
    |  Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie$^3$$^,$$^4$ , Emily J. Woodworth$^1$$^,$$^2$  
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$ 
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU
abstract:  |  
  1) DNA Barcoding has been remarkably successful in nearly all kingdoms of life and has allowed for the rapid analysis of ecological assemblages. Successful DNA barcoding in the plant kingdom has been more difficult than other kingdoms. Due to this understanding plants in ecological contexts and understandings of their synecology in some instances may begin to lag behind other kingdoms.  
  2) Here we utilize hyb-seq, museum studies, and species distribution modelling, to detect the plant species present in pollen loads collected from Queen Bumble Bees.  
  3) By utilizing Species distribution modelling we allow for one to process hyb-seq data, create user specified sequence databases which may use MORE ACCURATE alignment algorithms on personal computers over realistic time periods. 
  4) We show that hyb-seq using the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches in real world scenarios.  
  5) We conclude that these probes offer promise for the identification of plant tissue in both single sample, and metasample contexts.  
keywords: | 
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

```{r compare microscopic to molecular Pollen ID, eval = T}
a <- c('Trifolium', 'Lupinus', 'Glyccrrhiza', 'Mitella', 'Geum')
b <- c('Lupinus', 'Lathyrus', 'Potentilla', 'Androsace', 'Bistorta', 'Vicia')
c <- c('Jeffersonia', 'Micranthes', 'Prunus', 'Delphinium', 'Androsace', 
       'Penstemon', 'Orthocarpus', 'Scuttelaria', 'Aquilegia', 'Castilleja', 'Draba')
d <- c('Salix', 'Boechera')

@@

pmorph <- read.csv('../data/morpho_pollen.csv') 
blst <- read.csv('../data/blast_seqs_classified.csv') 
blst_seqs <- distinct(blst, sample.id)

@@

mertensia_samples <- pmorph %>% 
  filter(morphotype == 'MERTENSIA') %>% 
  select(sample.id, Prop, sum)

@@

blst_mertensia <- blst %>% 
  filter(Genus == 'Mertensia') %>% 
  select(sample.id,  Prcnt_seqs) %>% 
  group_by(sample.id) %>% 
  mutate(Prcnt_seqs = sum(Prcnt_seqs)) %>% 
  distinct(sample.id, .keep_all = T)

@@

# PERFORM FOR ERICACEAE TOO - A single record maybe come out?
ericaceae_samples <- pmorph %>% 
  filter(morphotype == 'ERICACEAE') %>% 
  select(sample.id, Prop, sum)

blst_ericoid <- filter(blst, Family == 'Ericaceae') # 44 might be Epilobium anagallidolium
ericaceae_detected <- ericaceae_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_ericoid$sample.id, 1, 0))

@@

blst_aster2 <- filter(blst_aster, Genus %in% 
                            c('Taraxacum', 'Helianthella', 'Scabrethia'))
aster2_detected <- aster2_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_aster2$sample.id, 1, 0))

@@

blst_aster1 <- filter(blst_aster, !Genus %in% c('Taraxacum', 'Helianthella', 'Scabrethia') )
aster1_detected <- aster1_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_aster1$sample.id, 1, 0))

@@

rm(mertensia_samples, blst, blst_aster, blst_aster1,
   blst_aster2, aster1_samples,  aster2_samples, a, b, c, d, 
   blst_mertensia, aster_samples, ericaceae_samples, blst_ericoid)
```

@@

>>>>

---
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging
  Patterns of Queen Bumble Bees
author:  |
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$ 
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU
    |  Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie$^3$$^,$$^4$ , Emily J. Woodworth$^1$$^,$$^2$  
abstract:  |  
  1) DNA Barcoding has been remarkably successful in nearly all kingdoms of life and has allowed for the rapid analysis of ecological assemblages. Successful DNA barcoding in the plant kingdom has been more difficult than other kingdoms. Due to this understanding plants in ecological contexts and understandings of their synecology in some instances may begin to lag behind other kingdoms.  
  2) Here we utilize hyb-seq, museum studies, and species distribution modelling, to detect the plant species present in pollen loads collected from Queen Bumble Bees.  
  3) By utilizing Species distribution modelling we allow for one to process hyb-seq data, create user specified sequence databases which may use MORE ACCURATE alignment algorithms on personal computers over realistic time periods. 
  4) We show that hyb-seq using the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches in real world scenarios.  
  5) We conclude that these probes offer promise for the identification of plant tissue in both single sample, and metasample contexts.  
keywords: | 
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  word_document: default
  toc: no
  pdf_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

```{r compare microscopic to molecular Pollen ID, eval = T}

@@

pmorph <- read.csv('../data/morpho_pollen.csv') 
blst <- read.csv('../data/blast_seqs_classified.csv') %>% 
  group_by(sample.id) %>% 
  mutate(Prcnt_Seqs_Class = sum(Prcnt_seqs)) %>% 
  ungroup()
blst_seqs <- distinct(blst, sample.id)

@@

mertensia_samples <- pmorph %>% 
  filter(morphotype == 'MERTENSIA') %>% 
  select(sample.id, morphotype, Prop, sum)

@@

blst_mertensia <- blst %>% 
  filter(Genus == 'Mertensia') %>% 
  select(sample.id,  Prcnt_seqs, Seqs_pr_taxon, Total_seqs, Prcnt_seqs, Prcnt_Seqs_Class) %>% 
  group_by(sample.id) %>% 
  mutate(Prcnt_seqs = sum(Prcnt_seqs),
         morphotype = 'MERTENSIA') %>% 
  distinct(sample.id, .keep_all = T)

@@

# PERFORM FOR ERICACEAE TOO - A single record maybe come out?
ericaceae_samples <- pmorph %>% 
  filter(morphotype == 'ERICACEAE') %>% 
  select(sample.id, Prop, morphotype)

blst_ericoid <- filter(blst, Family == 'Ericaceae') %>% # 44 might be Epilobium anagallidolium
  mutate(morphotype = 'ERICACEAE')
ericaceae_detected <- ericaceae_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_ericoid$sample.id, 1, 0))

@@

blst_aster2 <- filter(blst_aster, Genus %in% 
                            c('Taraxacum', 'Helianthella', 'Scabrethia')) %>% 
  mutate(morphotype = 'ASTERACEAE_2')
aster2_detected <- aster2_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_aster2$sample.id, 1, 0))

@@

blst_aster1 <- filter(blst_aster, !Genus %in% c('Taraxacum', 'Helianthella', 
                                                'Scabrethia') ) %>% 
  mutate(morphotype = 'ASTERACEAE_1')
aster1_detected <- aster1_samples %>% 
  mutate(DETECTED = if_else(sample.id %in% blst_aster1$sample.id, 1, 0))

@@

# Semi-quantitative accord between sequence % and Grain Count
palyn_cnt <- bind_rows(aster1_samples, aster2_samples, ericaceae_samples, mertensia_samples) %>% 
  select(morphotype, countProp = Prop)

pollen_props <- bind_rows(blst_aster1, blst_aster2, blst_mertensia, blst_ericoid) %>% 
  select(morphotype, sequenceProp = Prcnt_seqs, Seqs_pr_taxon, Total_seqs, Prcnt_Seqs_Class) %>% 
  left_join(., palyn_cnt, by = c('sample.id', 'morphotype')) %>% 
  drop_na()

rm(mertensia_samples, blst, blst_aster, blst_aster1,
   blst_aster2, aster1_samples,  aster2_samples, a, b, c, d, palyn_cnt,
   blst_mertensia, aster_samples, ericaceae_samples, blst_ericoid)
```

@@

QUANTITATIVE EVIDENCE !?!?!

```{r}

pol_p <- pollen_props %>% 
  mutate(sequenceProp = sequenceProp/100, 
         adjSeqProp = sequenceProp * (100/Prcnt_Seqs_Class))

summary(model <-  lm(sequenceProp ~ countProp, data = pol_p) ) # better with additive term + morphotype
summary(model)$adj.r.squared
summary(model)$p.val

newdat <- data.frame(
  countProp = seq(from = min(pol_p$countProp), to = max(pol_p$countProp), 
                  length.out = 100) )#, 
#  morphotype = sample(unique(pol_p$morphotype), replace = T, size = 100,
#       prob = as.numeric(table(pol_p$morphotype)) / nrow(pol_p)))

CI <- predict(model, newdat, se.fit = TRUE, interval = "confidence", level = 0.95)

par(pty="s")
plot(x = pol_p$countProp,  pol_p$sequenceProp, 
      xlim = c(0,1), ylim = c(0,1), axes = F, main = 'Correlation of Proportion Counted\nGrains and Sequence Reads',
     xlab = 'Grains Counted', ylab = 'Classified Sequences',
     pch=18, slty = 3, gap = 0.025)
axis(1)
axis(2)
abline(model)
lines(newdat$countProp, CI[['fit']][,3], lty = 'dotted', col = 'grey25')
lines(newdat$countProp, CI[['fit']][,2], lty = 'dotted', col = 'grey25')
text(0.2, 0.8, expression("adj. r"^2*" = " ))
text(0.4, 0.8, paste0(round(summary(model)$adj.r.squared, 3)))
text(0.3, 0.7, "p < 0.001")

rm(newdat)
```

```{r Run with correction for total sequences, eval = F}
## doesnt really make a difference - increase r2 by about 2% - with morphotype up to 6%, multiple r2 0.8

summary(model <-  lm(adjSeqProp ~ countProp, data = pol_p) )  #  + morphotype
newdat <- data.frame(
  countProp = seq(from = min(pol_p$countProp), to = max(pol_p$countProp), 
                  length.out = 100) )
CI <- predict(model, newdat, se.fit = TRUE, interval = "confidence", level = 0.95)

par(pty="s")
plot(x = pol_p$countProp,  pol_p$adjSeqProp, 
      xlim = c(0,1), ylim = c(0,1), axes = F, main = 'Correlation of Proportion Counted\nGrains and Sequence Reads',
     xlab = 'Grains Counted', ylab = 'Classified Sequences',
     pch=18, slty = 3, gap = 0.025)

axis(1)
axis(2)
abline(model)
lines(newdat$countProp, CI[['fit']][,3], lty = 'dotted', col = 'grey25')
lines(newdat$countProp, CI[['fit']][,2], lty = 'dotted', col = 'grey25')
text(0.2, 0.8, expression("adj. r"^2*" = " ))
text(0.4, 0.8, paste0(round(summary(model)$adj.r.squared, 3)))
text(0.3, 0.7, "p < 0.001")

rm(pol_p, newdat)
```

The relationship shows that grains which are often present in trace amounts have relatively high amounts of their abundances estimated, and grains which are present in high amounts have fewer sequences classified. This is considered to be due to the proportion of high false positives which occur in the classification process with NGS (BELL NOVEMEBER 2021)
